---
title: "A simple workflow using BOLDconnectR"
author: "Padhye SM, Ballesteros-Mejia CL,Agda TJA, Agda JRA, Ratnasingham S"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    highlight: pygments
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
    fig_width: 8
    fig_height: 5
    self_contained: true
vignette: >
  %\VignetteIndexEntry{BOLDconnectR Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# A simple workflow using BOLDconnectR

## Retrieval and Analysis of BOLD Data for the Fairy Shrimp Genus *Streptocephalus* (Crustacea: Anostraca)

The following vignette presents a workflow to demonstrate the main functionality of the **BOLDconnectR** package (i.e., `search`, `fetch` and `analyze`). 

Specifically, the workflow will show:

1.  Searching and fetching all public *Streptocephlaus* records available on BOLD in the BCDM format.
2.  Obtaining a concise summary of the downloaded data.
3.  Visualizing their occurrence distribution.
4.  Aligning the COI-5P sequences using `muscle` and visualizing the resulting `NJ` tree.
5.  Calculating BIN richness estimates and multisite beta diversity measure between different countries.

**BOLDconnectR** can be installed either by `install.packages` (for a stable CRAN version) or the devtools package (for the development version from GitHub)

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Installation and Import

```{r install,message=F,warning=F,echo=FALSE}
# install.packages("devtools")
# devtools::install_github("boldsystems-central/BOLDconnectR")
# install.packages('BOLDconnectR')
library(BOLDconnectR)
```

### Suggested packages

The sequence alignment and NJ tree functions require additional Bioconductor based packages to be installed and imported.

```{r suggests,message=F,warning=F,echo=FALSE,eval=F}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("msa", "Biostrings", "muscle"))

library(msa)
library(Biostrings)
library(muscle)
```

## API key

`bold.fetch()` requires an API key to retrieve the BCDM data.

```{r api-key, eval=T,include=FALSE}
# bold.apikey("")
```

#### Please note

The example dataset `test.data2` is used here for demonstrating the BOLDconnectR functionality due to the API key requirement for downloading data. The dataset represents a snapshot of public BOLD records obtained from the BOLD public data package released on June 30, 2026 (<https://doi.org/10.5883/DP-BOLD_Public.30-Jun-2026>). When reproducing this workflow at a later date using an API key, the query may return a different number of records because the underlying BOLD database is continuously updated.

## 1. Searching and fetching

`bold.public.search()` is used to retrieve public record IDs. `bold.full.search` can also be used instead in cases where user has private data stored. The `ids` returned using the `bold.public.search` are used in the `bold.fetch` to get the BCDM data. Here the search is also filtered based on the amplicon basepair number (`bold.fetch` *not used here as it requires an API key. A test dataset* `test.data2` *used instead*).

```{r fetch-data, message=F,warning=F,eval=T}
# Search
data_ids <- bold.public.search(taxonomy = list("Streptocephalus"))
# Fetch
streptocephalus_data <- bold.fetch(
  get_by = "processid",
  identifiers = data_ids$processid,
  filt_basecount = c(500, 670)
)
# streptocephalus_data <- data("test.data2")

DT::datatable((head(streptocephalus_data, 10)))
```

## 2. Concise summary of the data

A concise summary of the result is generated to get an idea of what all the data contains (*Note*: The `Amplicon_lenght_range` is now between 500-670)

```{r summarize-data, message=F,warning=F}
bcdm_summary <- bold.data.summarize(
  bold_df = streptocephalus_data,
  summary_type = "concise_summary"
)

DT::datatable(bcdm_summary$concise_summary)
```

## 3. Visualizing their occurrence distribution

```{r map-analysis, message=F,warning=F}
map_res <- bold.analyze.map(bold_df = streptocephalus_data)
```

## 4. Alignment and visualization

### 4a. Sequence alignment

A `muscle` algorithm is run on the *Streptocephalus* sequence data to get a multiple sequence alignment. The headers for each sequence are customized so as to contain `processid` and `bin_uri` of each sequence. **Please note:** the library `Biostrings`, `muscle` and `msa` need to be imported in the R session prior to sequence alignment.

```{r align-data, message=F,warning=F}
seq_align <- bold.analyze.align(
  bold_df = streptocephalus_data,
  marker = "COI-5P",
  cols_for_seq_names = c("processid", "bin_uri"),
  align_method = "Muscle"
)
DT::datatable(
  seq_align[, c("aligned_seq", "msa.seq.name")],
  options = list(pageLength = 10, scrollX = TRUE)
)
```

### 4b. Neighbor Joining Tree

The aligned sequences are then visualized via Neighbor Joining Tree. The output from `bold.analyze.align` is directly used for the visualization. **Please note:** the library `Biostrings`, `muscle` and `msa` need to be imported in the R session prior to sequence alignment

```{r tree-analysis,message=F,warning=F}
seq_tree <- bold.analyze.tree(
  bold_df = seq_align,
  dist_model = "K80",
  clus_method = "nj",
  tree_plot = TRUE,
  tree_plot_type = "p",
  save_dist_mat = TRUE,
  pairwise.deletion = TRUE
)

seq_tree$base_freq
```

## 5. Calculating BIN richness estimates and multisite beta diversity measure between different countries

### 5a. BIN richness

```{r diversity-analysis_alpha, message=F,warning=F}
diversity_res <- bold.analyze.diversity(
  bold_df = streptocephalus_data,
  taxon_rank = "species",
  site_type = "locations",
  location_type = "country.ocean",
  diversity_profile = "richness"
)
DT::datatable(diversity_res$richness)
```

### 5b. Beta diversity

```{r diversity-analysis_beta, message=F,warning=F}
beta_diversity_res <- bold.analyze.diversity(
  bold_df = streptocephalus_data,
  taxon_rank = "species",
  site_type = "locations",
  location_type = "country.ocean",
  diversity_profile = "beta",
  beta_index = "jaccard"
)

beta_diversity_res$total.beta
```

## Export data

Data can also be exported as a FASTA file, a multiple sequence alignment file (in FASTA format) or customized datasets of BCDM based on certain presets (e.g.,preset = taxonomy will export only taxonomy based information from all the BCDM data)

```{r export-data, eval=FALSE}
bold.export(
  bold_df = streptocephalus_data,
  export_type = "fas",
  cols_for_fas_names = c("bin_uri", "genus", "species"),
  export = "~/Desktop/boldconnectr_sequences.fas"
)
```

## Additional notes

In addition, several analysis functions return commonly used R data structures as part of their outputs, including an `occurrence matrix`, `sf` object, `DNABin` object containing unaligned sequences, and a `phylo` object generated from aligned sequences. These outputs can be readily integrated into workflows from other R packages for downstream analyses (e.g., using an occurrence matrix for ecological or biodiversity analyses with packages such as `vegan` and `betapart`).

For more/detailed information, please visit (<https://github.com/boldsystems-central/BOLDconnectR>) or use the `help` function in R.
